Fail Faster

### 2025-03-21 paper/paper.tex
wibble

Property-based testing (PBT) is a software testing technique that uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often given a very short time
window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
The faster a test fails, the better.
% The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
\iflater 
\bcp{Feels like we're missing a trick in the transition here.  Can we
  say, ``Because performance is important, PBT libraries are already
  quite tuned.  [Can we substantiate that??]  But more is
  possible...'' }
\fi

One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
that provide combinators for building and composing generators. However, careful measurements show that the performance of existing generator libraries falls far short of what
is possible.

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
the effectiveness of PBT.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we characterize the impact of \rand choice through a controlled experiment 
that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
combine high-level design with low-level performance.

To demonstrate the effectiveness and generality of our approach, which we call \name, we implement staged PBT libraries in multiple languages.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in replacements that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

@@

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
To assess the impact of our approach, we evaluate the performance and bug-finding speed of 
\name generators across a range of benchmarks. Our experiments compare generators 
built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with 
existing generator libraries (and their default randomness mechanisms). Specifically, 
we implement 
two fully semantically 
equivalent staged generator libraries that replicate the behavior of \bq 
in OCaml and ScalaCheck in Scala, allowing us to assess the effectiveness of our 
technique across different languages and runtime environments.

>>>>

Property-based testing (PBT) is a software-testing technique that uses
random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often given a very short time
window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
The faster it finds a failing test, the better.
% The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
\iflater
\bcp{Feels like we're missing a trick in the transition here.  Can we
  say, ``Because performance is important, PBT libraries are already
  quite tuned.  [Can we substantiate that??]  But more is
  possible...'' }
\fi

One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
that provide combinators for building and composing
generators. However, careful measurements show that the performance of
existing generator libraries---even well-engineered and widely used
ones---falls far short of what is possible.

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
be responsible for a surprisingly high proportion of its runtime. Combined, these factors limit
the effectiveness of PBT.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We address both sources of inefficiency.
First, we show how to use \textit{multi-stage programming}---a classic technique for
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of
many common generator abstractions.
Second, we characterize the impact of \rand choice through a controlled experiment 
that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.\bcp{We've done more than this, though: We've also
  demonstrated that you can get an X\% improvement in performance by
  using a specific RNG.}
Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
combine high-level design with low-level performance.

To demonstrate the effectiveness and generality of our approach\bcp{It
feels a bit grand to call what we've done an ``approach''---it's
really just a single performance-enhancing technique, staging, plus
the observation that the RNG matters and you should use the fastest
one you can.}, which we call \name, we implement staged PBT libraries in multiple languages.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in
replacements\bcp{that's not quite accurate, is it? I mean, they have
  the same functionality, but can you take a generator that works with
base\_quickcheck and just link it against our version without changing
the generator itself? That's what ``drop-in replacement'' means to me.} that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

@@

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
To assess the impact of our approach, we evaluate the performance and bug-finding speed of
\name generators across a range of benchmarks. Our experiments compare generators
built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with
existing generator libraries (and their default randomness mechanisms). Specifically,
we implement
two fully semantically
equivalent staged generator libraries that replicate the behavior of \bq
in OCaml and ScalaCheck in Scala, allowing us to assess the effectiveness of our
technique across different languages and runtime
environments.\bcp{This says ``assess the impact'' twice...}
